---
title: "Getting Started with RESOLVE"
author: "Gilles Colling"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with RESOLVE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# The code below is illustrative: it shows the current API but is not executed
# when the vignette is built (it needs the prebuilt resolve_c library and is
# covered by the test suite). `purl = FALSE` keeps the code-running check from
# tangling and executing it.
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  purl = FALSE
)
```

## Introduction

RESOLVE predicts sample-level outcomes from compositional data: species lists,
item sets, or any variable-length categorical input. It encodes the composition
into a learned representation and trains a multi-task model on top of it. The R
package is a thin client over the C++ engine.

## Installation

The R package itself is small and has no heavy dependencies. Install it, then
fetch the `resolve_c` backend (the C ABI over the engine, which bundles
libtorch) once with `resolve.install_backend()` -- the same delayed-binary model
the `torch` package uses:

```{r install}
remotes::install_github("gcol33/resolve", subdir = "r")

library(resolve)
resolve.install_backend()          # downloads the CPU backend on first use
# resolve.install_backend(variant = "cuda")   # GPU build, where published

resolve.available()                # TRUE once the backend is loaded
```

Alternatively, point the `RESOLVE_C_HOME` environment variable at a directory
that already contains a built `resolve_c` library (for example one built from
the monorepo `src/core`) and it is loaded automatically at package load.

## Quick Start

RESOLVE reads two CSV files:

1. **Header file**: one row per sample (plot) with an ID, optional coordinates
   and covariates, and the target columns.
2. **Species file**: one row per species-in-sample occurrence with a sample ID,
   a species ID, and an optional abundance.

### Load a dataset

```{r load-data}
library(resolve)

dataset <- resolve.dataset.csv(
  header  = "header.csv",
  species = "species.csv",
  roles = list(
    plot_id    = "plot_id",
    species_id = "species_id",
    abundance  = "cover",
    longitude  = "lon",
    latitude   = "lat",
    genus      = "genus",
    family     = "family"
  ),
  targets = list(
    area = list(column = "area", task = "regression", transform = "log1p")
  ),
  config = list(
    species_encoding = "hash",
    hash_dim = 32,
    top_k = 5
  )
)

# Inspect the inferred schema
str(dataset$schema())
```

### Train a model

`resolve.train.dataset()` builds a model from the dataset's schema, trains it,
and returns the trainer, the training result, and the dataset.

```{r train}
fit <- resolve.train.dataset(
  dataset,
  hiddenDims = c(256, 128, 64),
  maxEpochs  = 200,
  patience   = 25,
  batchSize  = 4096,
  device     = "cpu",      # "cuda" is available via the GPU build of resolve_c
  lossConfig = "mae",
  verbose    = TRUE
)

fit$result$final_metrics

# Persist the trained model
resolve.save(fit$trainer, "model.pt")
```

### Predict

```{r predict}
predictor <- resolve.load("model.pt")

# Encode new data against the same vocabularies the model was trained on
new_dataset <- resolve.dataset.csv(
  header  = "new_header.csv",
  species = "new_species.csv",
  roles   = list(plot_id = "plot_id", species_id = "species_id"),
  targets = list(area = list(column = "area", task = "regression")),
  schemaSource = dataset
)

preds <- resolve.predict.dataset(predictor, new_dataset)
head(preds$predictions$area)
```

## Species encoding modes

The `species_encoding` config field selects how a sample's species set is
turned into features:

```{r encoding-modes}
# Feature hashing (default): fixed width, handles unseen species
config_hash <- list(species_encoding = "hash", hash_dim = 64)

# Learned per-species embeddings of the top-k species
config_embed <- list(species_encoding = "embed", top_k_species = 20)

# Rank-pooled species embeddings with weighted pooling
config_pool <- list(species_encoding = "rank_pool", pool_weighting = "log1p")
```

## Multi-task learning

Pass several targets to predict them jointly:

```{r multi-task}
targets <- list(
  area = list(column = "area", task = "regression", transform = "log1p"),
  habitat = list(column = "habitat_class", task = "classification")
)
```

## Loss configurations

```{r loss-config}
# Pure MAE (default, stable)
fit <- resolve.train.dataset(dataset, lossConfig = "mae")

# SMAPE (scale-invariant)
fit <- resolve.train.dataset(dataset, lossConfig = "smape")

# Phased: MAE -> MAE + SMAPE -> MAE + SMAPE + band penalty
fit <- resolve.train.dataset(dataset, lossConfig = "combined")
```

## Metrics

The package exposes the engine's metric helpers for evaluating predictions:

```{r metrics}
resolve:::resolve_mae(pred, actual)
resolve:::resolve_rmse(pred, actual)
resolve:::resolve_smape(pred, actual)
```

## Session info

```{r session-info, eval = TRUE, purl = FALSE}
sessionInfo()
```
