---
title: "Getting Started with RESOLVE"
author: "Gilles Colling"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with RESOLVE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

RESOLVE (**R**epresentation **E**ncoding of **S**pecies **O**utcomes via **L**inear **V**ector **E**mbeddings) predicts plot-level environmental attributes from species composition data. It treats species composition as *biotic context*—a rich, structured signal that encodes information about plot attributes.

## Installation

```{r install}
# Install from GitHub
remotes::install_github("gcol33/resolve", subdir = "r")
```

## Quick Start

### Prepare Your Data

RESOLVE requires two CSV files:

1. **Header file**: One row per plot with plot ID, coordinates, covariates, and target variables
2. **Species file**: One row per species-plot occurrence with species ID, plot ID, and abundance

```{r data-structure}
# Example header file structure
header <- data.frame(
  plot_id = c("P001", "P002", "P003"),
  longitude = c(4.35, 4.36, 4.37),
  latitude = c(50.85, 50.86, 50.87),
  elevation = c(100, 120, 95),
  biomass = c(45.2, 52.1, 38.7)
)

# Example species file structure
species <- data.frame(
  plot_id = c("P001", "P001", "P002", "P002", "P003"),
  species = c("Quercus robur", "Fagus sylvatica", "Quercus robur",
              "Pinus sylvestris", "Fagus sylvatica"),
  cover = c(40, 25, 35, 45, 60),
  genus = c("Quercus", "Fagus", "Quercus", "Pinus", "Fagus"),
  family = c("Fagaceae", "Fagaceae", "Fagaceae", "Pinaceae", "Fagaceae")
)
```

### Load and Configure Dataset

```{r load-data}
library(resolve)

# Define column roles
roles <- list(
  plotId = "plot_id",
  speciesId = "species",
  speciesPlotId = "plot_id",
  abundance = "cover",
  genus = "genus",
  family = "family",
  x = "longitude",
  y = "latitude"
)

# Define prediction targets
targets <- list(
  biomass = list(
    column = "biomass",
    task = "regression",
    transform = "log1p"
  )
)

# Create dataset
dataset <- resolve.dataset(
  header = "path/to/header.csv",
  species = "path/to/species.csv",
  roles = roles,
  targets = targets,
  hashDim = 32,
  topK = 5
)

# Check dataset schema
print(dataset$schema)
```

### Train a Model

```{r train}
# Train with default parameters
result <- resolve.train(
  dataset,
  maxEpochs = 500,
  patience = 50,
  batchSize = 4096,
  device = "cpu",  # Use "cuda" for GPU
  verbose = TRUE
)

# View training results
print(result$result$final_metrics)

# Save the model
resolve.save(result, "my_model.pt")
```

### Make Predictions

```{r predict}
# Load a saved model
predictor <- resolve.load("my_model.pt")

# Prepare new data
new_dataset <- resolve.dataset(
  header = "path/to/new_header.csv",
  species = "path/to/new_species.csv",
  roles = roles,
  targets = targets
)

# Generate predictions
predictions <- resolve.predict(
  predictor,
  new_dataset,
  outputSpace = "raw"  # Inverse-transform to original scale
)

# View predictions
head(predictions$biomass)
```

## Species Encoding Options

RESOLVE supports two species encoding strategies:

### Hash Encoding (Default)

Feature hashing creates a fixed-dimension representation regardless of species vocabulary size. This is memory-efficient and handles unseen species gracefully.

```{r hash-encoding}
result <- resolve.train(
  dataset,
  speciesEncoding = "hash",
  hashDim = 64  # Increase for larger vocabularies
)
```

### Learned Embeddings

For datasets with consistent species vocabularies, learned embeddings can capture species-specific patterns.

```{r embed-encoding}
result <- resolve.train(
  dataset,
  speciesEncoding = "embed"
)
```

## Confidence-Based Predictions

RESOLVE tracks the fraction of unknown species per plot. Use confidence thresholds to filter unreliable predictions:

```{r confidence}
# Only return predictions for plots with >50% known species
predictions <- resolve.predict(
  predictor,
  new_dataset,
  confidenceThreshold = 0.5
)

# Predictions below threshold are set to NA
sum(is.na(predictions$biomass))
```

## Multi-Task Learning

RESOLVE supports simultaneous prediction of multiple targets:

```{r multi-task}
targets <- list(
  biomass = list(
    column = "biomass",
    task = "regression",
    transform = "log1p"
  ),
  habitat = list(
    column = "habitat_class",
    task = "classification"
  )
)

dataset <- resolve.dataset(
  header = "header.csv",
  species = "species.csv",
  roles = roles,
  targets = targets
)

result <- resolve.train(dataset)
```

## Loss Configuration

Three loss configurations are available for regression tasks:

```{r loss-config}
# Pure MAE loss (default, stable training)
result <- resolve.train(dataset, lossConfig = "mae")

# SMAPE loss (scale-invariant)
result <- resolve.train(dataset, lossConfig = "smape")

# Phased training: MAE -> MAE+SMAPE -> MAE+SMAPE+band penalty
result <- resolve.train(dataset, lossConfig = "combined")
```

## GPU Acceleration

For large datasets, enable CUDA acceleration:

```{r gpu}
result <- resolve.train(
  dataset,
  device = "cuda",
  batchSize = 8192  # Larger batches for GPU
)
```

## Session Info

```{r session-info, eval=TRUE}
sessionInfo()
```
